---
layout: post
title:  "Does R understand physical quantities?"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
comments: true
author: Edzer Pebesma
categories: r
---
<script src="https://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML" type="text/javascript"></script>

TOC

[DOWNLOADHERE]

### Stevens's measurement scales 

S.S. Stevens's classical 1946
[paper](http://psychology.okstate.edu/faculty/jgrice/psyc3214/Stevens_FourScales_1946.pdf)
_On the Theory of Scales of Measurement_ tells us there are four
measurement scales:

* nominal,
* ordinal,
* interval, and
* ratio.

R is pretty good at representing the first two by using `factor` and
`ordered`, 
```{r}
(f = factor(c("d", "a", "b", "c", "a", "b")))
(o = ordered(c("d", "a", "b", "c", "a", "b")))
```
which give warnings about meaningless operations, like
```{r}
(e = f * 2)
```
and R combines interval and ratio into `numeric` variables. Having
different representations between these different measurement scales
has, in my opinion, always been a major advantage of R. It prevents you
from doing things that are  _statistically_ not meaningful.

### Why physical units?

In physics class, we learned that every physical quantity has a
[measurement unit](https://en.wikipedia.org/wiki/Units_of_measurement).
If `a` represents speed, with unit `m/s`, we can't add it
meaningfully to `b` which has unit seconds, but we can add it to `c`
measured in `km/h` after proper unit conversion.  [Dimensional
analysis](https://en.wikipedia.org/wiki/Dimensional_analysis)
tracks units of measurements of variables when computations are
performed. It is used to determine the unit of measure of the result,
but also catches computations that aren't  _physically_ meaningful.
Can this be automated?

### Physical unit databases, and conversion software

[The Unified Code for Units of Measure](http://unitsofmeasure.org/trac), or UCUM,
is based on the _ISO 80000: 2009 Quantities and Units_ standards series
that specify the use of System International (SI). UCUM comes with
a BNF grammar and a machine-readable (XML) document with all the
units, or all those that are useful -- the amount of derivable
units is infinite.

Being rather formal, and close to ISO, it is no surprise that UCUM
has been recommended for encoding units of measures by many [open
geospatial consortium](http://www.opengeospatial.org/) standards
for spatial data.

A more pragmatic and practical approach is taken
by [udunits](https://www.unidata.ucar.edu/software/udunits/),
developed by the geo/atmospheric scientists of
[UCAR/unidata](https://www.unidata.ucar.edu/). Udunits not only
consists of an XML file with all the units, their names and
symbols, but also of a software library that can validate units,
check whether they are convertible (like km/h and m/s) and _carry
out this conversion_.  James Hiebert wrote an R package,
[udunits2](https://cran.r-project.org/package=udunits2), which
interfaces to this software library, but does little more than
exposing its functions as R functions.

### Using physical units in R: the units package

I have always wondered why R has no support for dimensions built
in, or at least have a package that does this. `Date` and `POSIXt`
objects have implicit units (1 day, 1 second), but only time 
difference `difftime` objects have explicit, and modifiable units:
```{r}
t = Sys.time() + 0:3 * 3600
(deltat = diff(t))
units(deltat) = "mins"
deltat
```

When I discovered the
[udunits2](https://cran.r-project.org/package=udunits2)
R package, I couldn't resist writing the
[units](https://cran.r-project.org/package=units) R package, which
works similarly to `difftime`, but for all physical units supported
by udunits2. Thus, after
```{r}
library(units)
(a = as.units(1:5, "m/s"))
```
we can do simple arithmetic:
```{r}
2 * a
a + a
a * a
```
but also automatic unit conversion
```{r}
b = as.units(1:5, "km/h")
a + b
b + a
a / b
a * b
```
as you can see, units are propagated and converted to that of
the first argument when needed, but not simplified. Wrong units trigger
an error:
```{r}
s = as.units(1:5, "s")
e = try(x <- a + s)
attr(e, "condition")[[1]]
```
We can also do comparison and apply basic functions, subset, or concatenate
```{r}
signif(a^2.5, 3)
a[2:4]
c(a,b)
c(b,a)
```
Conversion to and from `difftime`, use in `data.frame`s
or `matrix` objects is illustrated in the package
[vignette](https://cran.r-project.org/web/packages/units/vignettes/units.html).


## The further future

When dealing with measurement unit rigorously, the
output of linear regression of two variables, `zinc` with units
`ppm` and `dist` with units `m` would ideally contain them:
```
> library(sp)
> data(meuse)
> summary(lm(zinc ~ dist, meuse))

Call:
lm(formula = zinc ~ dist, data = meuse)

Residuals (ppm):
    Min      1Q  Median      3Q     Max 
-475.20 -189.94  -52.94  120.15 1088.80 

Coefficients:
            Estimate Units Std. Error t value Pr(>|t|)    
(Intercept)   756.70   ppm      35.66   21.22   <2e-16 ***
dist        -1195.67 ppm/m     114.84  -10.41   <2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 281.7 ppm on 153 degrees of freedom
Multiple R-squared:  0.4147,	Adjusted R-squared:  0.4109 
F-statistic: 108.4 on 1 and 153 DF,  p-value: < 2.2e-16
```
I'm convinced this would help understand what residuals, regression
coefficient estimates, and standard errors mean.

Getting output like this automatically may not happen any time soon:
when solving the normal equations, each entry of the cross product
matrix $X'X$ would need to store its own physical unit, and matrix
product and solve routines would need to propagate them.

## The near future

It is of course good to know whether R variables are stored
as `factor` or `character`, as `integer` or `double`, but it
doesn't prevent you from adding apples and oranges. Verifying
compatibility of physical units does.  [Dimensional
analysis](https://en.wikipedia.org/wiki/Dimensional_analysis)
helps here, and helps understanding and verifying meaningfulness
of results.

I would be more than happy to hear of any use cases using the units
package, be it for educational or operational projects, and also
for suggestions how (or pull requests) to improve this package. My
wish list right now:

* add units by default to axis labels of plots
* support log-units handling of udunits
* integrate with spatial and temporal reference systems in R
* link this to our work on [meaningful spatial statistics](http://www.sciencedirect.com/science/article/pii/S1364815213001977) and provenance of [data generation](http://ifgi.uni-muenster.de/~epebe_01/generativealgebra.pdf)

